
# LAGOS Spatial Analysis


```{r setup-spatial, include=FALSE}
library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
library("gridExtra")
```




## Loading in data


### First download and then specifically grab the locus (or site lat longs)

```{r LagoDownload}
#Lagos download script
#LAGOSNE::lagosne_get(dest_folder = LAGOSNE:::lagos_path())

#Load in lagos
lagos <- lagosne_load()
#Grab the lake centroid info
lake_centers <- lagos$locus


```



### Convert to spatial data
```{r LagoSpatial}
#Look at the column names
#names(lake_centers)

#Look at the structure
#str(lake_centers)

#View the full dataset
#View(lake_centers %>% slice(1:100))

spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'), crs=4326) %>%
  st_transform(2163)

#Subset for plotting
subset_spatial <- spatial_lakes %>%
  slice(1:100) 

subset_baser <- spatial_lakes[1:100,]

#Dynamic mapviewer
mapview(subset_spatial)

```


### Subset to only Minnesota

```{r LagoMN, fig.cap="Lakes in Minnesota by size"}
states <- us_states()

#Plot all the states to check if they loaded
#mapview(states)
minnesota <- states %>%
  filter(name == 'Minnesota') %>%
  st_transform(2163)


#Subset lakes based on spatial position
minnesota_lakes <- spatial_lakes[minnesota,]

#Plotting the first 1000 lakes
minnesota_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha') + mapview(minnesota)
```



## Map of Iowa & Illinois

```{r LagoIAIL, fig.cap="Outline of Iowa & Illinois"}
#Plot all the states to check if they loaded
#mapview(states)
iowa <- states %>%
  filter(name == 'Iowa') %>%
  st_transform(2163)

illinois <- states %>%
  filter(name == 'Illinois') %>%
  st_transform(2163)

mapview(illinois, alpha.regions = 0.4, aplha = 1, col.regions = "yellow") + 
mapview(iowa, alpha.regions = 0.4, aplha = 1, col.regions = "red")

iowa_lakes <- spatial_lakes[iowa,]
illinois_lakes <- spatial_lakes[illinois,]


```


```{r LagoIAILLakes, eval= FALSE, echo= FALSE}

#Plotting the first 1000 lakes
mapview(iowa, alpha.regions = 0.4, aplha = 1, col.regions = "red") +
  iowa_lakes %>%
  arrange(-lake_area_ha) %>%
  slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha') +
mapview(illinois, alpha.regions = 0.4, aplha = 1, col.regions = "yellow") +
  illinois_lakes %>%
  arrange(-lake_area_ha) %>%
  slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha')
```



## Subset LAGOS data

Subset LAGOS data to these sites, how many sites are in Illinois and Iowa combined? How does this compare to Minnesota?

```{r LagoSubset}
minnesota_length = length(minnesota_lakes$lagoslakeid)
iowa_illinois_lakes = rbind(illinois_lakes, iowa_lakes)
iowa_illinois_length = length(iowa_illinois_lakes$lagoslakeid)

```


Minnesota has `r minnesota_length` lakes while Iowa & Illinois combined have `r iowa_illinois_length` lakes


## Distribution of lake size in Iowa vs. Minnesota

Minnesota has bigger lakes than Iowa. Both are skewed to the right.

```{r LagoHist, echo= FALSE, fig.cap="Distribution of lake size in Iowa vs. Minnesota"}
grid.arrange(
ggplot(minnesota_lakes, aes(x= lake_area_ha)) + 
  geom_histogram(bins=10) + 
  scale_x_log10(labels = scales::comma) +
  labs(title="Minnesota Lakes") +
  xlab("Lake Area (HA)") +
  ylab("Count")
,
ggplot(iowa_lakes, aes(x= lake_area_ha)) + 
  geom_histogram(bins=10) + 
  scale_x_log10(labels = scales::comma) +
  labs(title="Iowa Lakes") +
  xlab("Lake Area (HA)") +
  ylab("Count")
  
)



```

## Lakes in Iowa & Illinois by lake area

```{r LagoIAILArea, fig.cap='Lakes in Iowa & Illinois by lake area'}

iowa_illinois_map = iowa_illinois_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha',  canvas = TRUE, at=c(0,10,100,1000,10000)) 

mapview(illinois,  canvas = TRUE , alpha.regions = 0.4, aplha = 1, col.regions = "yellow") + 
mapview(iowa,  canvas = TRUE , alpha.regions = 0.4, aplha = 1, col.regions = "red") +
iowa_illinois_map 

```


##  Future investigation

What other data sources might we use to understand how reservoirs and natural lakes vary in size in these three states? 

Ground water & especially water table depth data would be helpful for looking at how the lakes interact with one another. Long term weather & climate (rainfall, snowfall, and Total Solar Radiance) would be helpful in to for understanding water origin & evaporation.





