# Weather and Agricultural Yield Regressions

By looking at weather patterns we can gleam new insight into farm yields. Here we use [PRISM temperture](https://prism.oregonstate.edu/) data and [NASS](https://www.nass.usda.gov/Data_and_Statistics/) corn & soybean yield data. 

```{r CornSetup, include=FALSE}
library(tidyverse)
library(R.matlab)
library(rnassqs)
library(gridExtra)
library(mapview)
library(USAboundaries)
library(USAboundariesData)
library(leaflet.extras2)
library(raster)

```

## Counties

Get all counties in Iowa

```{r}
counties = us_counties(map_date = NULL, resolution = c("low", "high"), states = 'iowa') %>%
  dplyr::select(name, geometry)
counties$name = tolower(counties$name)

```

## Temperture

Load the PRISM daily maximum temperatures for Iowa

```{r TempRegression, fig.cap='Daily Maximum Temperature, Iowa County #1', warning=FALSE}

# daily max temperature
# dimensions: counties x days x years
prism <- readMat("data/corn/prismiowa.mat")

# view data for county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
#t_1981_c1[366]

ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("Day of Year") +
  ylab("Daily Maximum Temperature (°C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")


```
```{r TempTidy}

# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)

```

### Summer temperature trends: Winneshiek County

```{r TempSummerWin, fig.cap='Summer temperature trends: Winneshiek County', echo=FALSE, message=FALSE}

tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

winneSummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meanTmax = mean(tmax))

ggplot(winneSummer, mapping = aes(x = year, y = meanTmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "Year", y = "Tmax (°C)", title='Summer temperature trends: Winneshiek County') +
  geom_smooth(method = lm)

lm_summerTmax <- lm(meanTmax ~ year, winneSummer)

summary(lm_summerTmax)

```

### Winter temperature trends: Winneshiek County

```{r TempWinterWin, fig.cap='Winter temperature trends: Winneshiek County', echo=FALSE, message=FALSE}

winneWinter <- tmaxdf %>%
  filter(countyfp==191 & (doy <= 59 | doy >= 335) & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meanTmax = mean(tmax))

ggplot(winneWinter, mapping = aes(x = year, y = meanTmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "Year", y = "Tmax (°C)", title='Winter temperature trends: Winneshiek County') +
  geom_smooth(method = lm)

lm_winterTmax <- lm(meanTmax ~ year, winneWinter)
summary(lm_winterTmax)

```

### Multiple Regression - Quadratic Time Trend

```{r TempQuad, fig.cap='Multiple Regression - Quadratic Time Trend', echo=FALSE}

winneWinter$yearSq <- as.numeric(winneWinter$year)^2

lm_winterTmaxQuad <- lm(meanTmax ~ year + yearSq, winneWinter)
winneWinter$fitted <- lm_winterTmaxQuad$fitted.values

ggplot(winneWinter) +
  geom_point(mapping = aes(x = year, y = meanTmax)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "Year", y = "Tmax (°C)", title='Multiple Regression - Quadratic Time Trend')

summary(lm_winterTmaxQuad)


```


## Corn

Download NASS corn yield data

```{r CornDownload}

# set our API key with NASS. Stored in the env to prevent publishing on GitHub
# You can get your own key by visiting:
# https://quickstats.nass.usda.gov/api
nassqs_key <- Sys.getenv("nassqs_auth")
nassqs_auth(key = nassqs_key)

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")
#cornyieldsall <- nassqs_yields(params)
```
```{r CornDownload2, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, include=FALSE, cache=FALSE}
cornyieldsall <- nassqs_yields(params)
```
```{r CornDownload3}


# clean and filter this dataset
cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$year <- as.numeric(cornyieldsall$year)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)
cornyields <- dplyr::select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields$county_name = tolower(cornyields$county_name)
cornyields <- tibble(cornyields)

```

### Yield over time

Corn yield has consistently increased year over year. For the linear model, with every increasing year we would expect an increase in yield of 2.457 bushels / acre. This agrees with my prior knowledge from working at Land O'Lakes. Improvements in farming practices and seed genetics have been driving this trend. Although adoption of these changes varies from farm to farm, at the aggregated county level this affect can be clearly seen in the data. 

```{r, fig.cap='Yield over time', res=600, echo=FALSE, message=FALSE, warning=FALSE}


winneYield = cornyields %>%
  filter(county_ansi==191)

ggplot(winneYield, mapping = aes(x = year, y = yield)) +
  geom_point( alpha = 0.4, colour = "black") +
  theme_bw() +
  labs(x = "Year", y = "Yield (bsh/ac)", title= "Yield over time", subtitle="Winneshiek, IA") +
  geom_smooth(method = lm)

```


### Quadratic time trend

There is no evidence of slowing corn yield growth in Winneshiek County, IA. The slope of the quadratic time model is relatively constant over time, if anything there is slight acceleration. 

```{r CornTimeQuad, fig.cap='Corn Quadratic Time Trend', res=600, message=FALSE}
winneYieldLin<- lm(yield ~ year, winneYield)
#summary(winneYieldLin)
winneYield$lin <- winneYieldLin$fitted.values


winneYield$yearSq <- as.numeric(winneYield$year)^2
winneYieldQuad<- lm(yield ~ year + yearSq, winneYield)
#summary(winneYieldQuad)
winneYield$quad <- winneYieldQuad$fitted.values

ggplot(winneYield) +
  geom_point(mapping = aes(x = year, y = yield), alpha = 0.4, colour = "black") +
  geom_line(mapping = aes(x = year, y = quad, col='Quadratic Year'), size=1) +
  geom_line(mapping = aes(x = year, y = lin, col='Linear Year'), size=1) +
  theme_bw() +
  labs(x = "Year", y = "Yield (bsh/ac)", title= "Yield over time", subtitle="Winneshiek, IA", color = "Model")

```

### Time Series

The yield peaks at a mean summer temperature max of 27 Celsius according to the quadratic temp model.This follows intuition that plants respond well to increasing temperature, if it was freezing they would not grow. However, increased temperature only increases yield up to 27 degrees than than gradually drops. The quadratic temperature model has significantly greater predictive performance than the linear. However with such a significant year over year affect, the Quadratic Temp Year model has the best performance. We can see that Year Squared has a significantly bigger affect on Yield than the Temperature does as shown by the p-values of the coefficients. By incorporating year quared & mean temperture max squared, we account for 81% of the variation observed in yield (Adjusted R-squared).


```{r CornTimeSeries, message=FALSE}
# add squares for quadratic modeling
winneYieldTemp = merge(winneYield, winneSummer, by = 'year') %>%
  mutate(
    yieldSq = yield^2,
    meanTmaxSq = meanTmax^2
  )
# Linear model of Tmax on Yield
winneYieldTempLin<- lm(yield ~ meanTmax, winneYieldTemp)
winneYieldTemp$tempLin <- winneYieldTempLin$fitted.values

# Quadratic model of year on yield
winneYieldTempQuad<- lm(yield ~ meanTmax + meanTmaxSq, winneYieldTemp)
winneYieldTemp$tempQuad <- winneYieldTempQuad$fitted.values

# Linear model of year & Tmax on yield
winneYieldYearLinTempLin<- lm(yield ~ meanTmax + year, winneYieldTemp)
winneYieldTemp$yearTempLin <- winneYieldYearLinTempLin$fitted.values

# Quadratic model of year & Tmax on yield
winneYieldYearTempQuad<- lm(yield ~ year + yearSq + meanTmax + meanTmaxSq, winneYieldTemp)
winneYieldTemp$yearTempQuad <- winneYieldYearTempQuad$fitted.values
#summary(winneYieldYearTempQuad$coefficients)

```
```{r, fig.cap='Yield models', res=600, echo=FALSE, warning=FALSE, message=FALSE}

plot_win_temp =  ggplot(winneYieldTemp) +
  geom_point(mapping = aes(x = meanTmax, y = yield), alpha = 1, colour = "black") +
  geom_line(mapping = aes(x = meanTmax, y = tempQuad, col="Quadratic Temp"), size=1) +
  geom_line(mapping = aes(x = meanTmax, y = tempLin, col="Linear Temp"), size=1) +
  geom_line(mapping = aes(x = meanTmax, y = yearTempLin, col="Linear Temp Year"), size=1) +
  geom_line(mapping = aes(x = meanTmax, y = yearTempQuad, col="Quadratic Temp Year "), size=1) +
  theme_bw() +
  labs(x = "T Max (°C)", y = "Yield (bsh/ac)", title= "Yield over T Max", subtitle="Winneshiek, IA", color = "Model")

plot_win_yield=  ggplot(winneYieldTemp) +
  geom_point(mapping = aes(x = year, y = yield), alpha = 1, colour = "black") +
  geom_line(mapping = aes(x = year, y = tempQuad, col="Quadratic Temp"), size=1) +
  geom_line(mapping = aes(x = year, y = tempLin, col="Linear Temp"), size=1) +
  geom_line(mapping = aes(x = year, y = yearTempLin, col="Linear Temp & Year"), size=1) +
  geom_line(mapping = aes(x = year, y = yearTempQuad, col="Quadratic Temp & Year "), size=1) +
  theme_bw() +
  labs(x = "Year", y = "Yield (bsh/ac)", title= "Yield over time", subtitle="Winneshiek, IA", color = "Model") 

grid.arrange(
  plot_win_temp,
  plot_win_yield  
)
```

```{r, include=FALSE}
summary(winneYieldYearTempQuad)

```





### Time Cross-Section

We see the same relationship between Summer Mean T Max and Yield in 2018 for all counties in Iowa that we saw for all years in Winneshiek. The peak is at 28 degrees instead of 27 degrees. This data tells the story better, as unlike the Winneshiek analysis, this is not affected by year of year yield increases due to improved farming methods. Looking at only one year the p-value of the mean temperture mac goes from  0.0005 to 0.00239 when we only look at one year of data.

```{r}
tmaxAll = tmaxdf %>%
  filter(doy >= 152 & doy <= 243) %>%
  group_by(year, countyfp) %>%
  summarize(meanTmax = mean(tmax))

iowa = merge(cornyields, tmaxAll, by.x=c("year", "county_ansi"), by.y=c("year", "countyfp")) %>% 
  mutate(
    yearSq = as.numeric(year)^2,
    yieldSq = yield^2,
    meanTmaxSq = meanTmax ^ 2
  )

iowa2018 = iowa %>%
  filter(year == 2018)

```


```{r}
iowa2018YieldTempLin<- lm(yield ~ meanTmax, iowa2018)
#summary(iowa2018YieldTempLin)
iowa2018$lin <- iowa2018YieldTempLin$fitted.values

iowa2018YieldTempQuad<- lm(yield ~ meanTmax + meanTmaxSq, iowa2018)
#summary(iowa2018YieldTempQuad)
iowa2018$quad <- iowa2018YieldTempQuad$fitted.values


```

```{r, fig.cap='Corn Yield over T Max for Iowa in 2018', fig.width = 8, res=600, echo=FALSE}
ggplot(iowa2018) +
  geom_point(mapping = aes(x = meanTmax, y = yield), alpha = 0.4, colour = "black") +
  geom_line(mapping = aes(x = meanTmax, y = quad, colour="Quadratic"), size=2) +
  geom_line(mapping = aes(x = meanTmax, y = lin, colour="Linear"), size=2) +
  theme_bw() +
  labs(x = "T Max (°C)", y = "Yield (bsh/ac)", title= "Yield over T Max", subtitle="Iowa 2018", color = "Model")

```



### Panel Regession

One way to leverage multiple time series is to group all data into what is called a "panel" regression. Combined the yield models tell a more complete picture with an Adjusted R-squared of 0.7225 . On its own, the quadratic temperature model only have predictive power up to 150 bushels, after which all increases are dependent on factors outside of temperature. The year model provides the best single factor fit, following the slope most tightly. The county also plays a large factor in that a 50 bushel difference can be attributed to the location of your farm. Some counties had a greater signficance like CLARKE but this was not true for most counties.


```{r}
# add factors
iowa$county_name = as.factor(iowa$county_name)
iowa$county_ansi = as.factor(iowa$county_ansi)

# Single Factor Models
iowa_TempQuad = lm(yield ~ meanTmax + meanTmaxSq, iowa)
#summary(iowa_TempQuad)
iowa$m_tempSq <- iowa_TempQuad$fitted.values

iowa_YearQuad = lm(yield ~ year + yearSq, iowa)
#summary(iowa_YearQuad)
iowa$m_yearSq <- iowa_YearQuad$fitted.values

iowa_County = lm(yield ~ county_name, iowa)
#summary(iowa_County)
iowa$m_county <- iowa_County$fitted.values

iowa_Combined = lm(yield ~ year + yearSq + meanTmax + meanTmaxSq + county_name, iowa)
#summary(iowa_Combined)
iowa$m_combined <- iowa_Combined$fitted.values

#summary(iowa_Combined)


```

```{r, fig.cap="Single Factor Models vs Actual Corn Yield in Iowa for all years", fig.width = 8, fig.height= 12, res=600, echo=FALSE}
grid.arrange(
top = "Single Factor Models vs Actual Corn Yield in Iowa for all years",
layout_matrix = rbind(c(1, 2, 3),c(4, 4, 4)),

ggplot(iowa) +
  geom_point(mapping = aes(x = yield, y = m_tempSq), alpha = 0.2) +
  geom_line(mapping = aes(x = yield, y = yield), linetype = "dashed", size=1.5) +
  theme_bw() +
  labs(x = "Actual Yield", y = "Fitted Yield", title= "Quadratic Temp",)
,
ggplot(iowa) +
  geom_point(mapping = aes(x = yield, y = m_yearSq), alpha = 0.2) + 
  geom_line(mapping = aes(x = yield, y = yield), linetype = "dashed", size=1.5) +
  theme_bw() +
  labs(x = "Actual Yield", y = "Fitted Yield", title= "Quadratic Year")
,
ggplot(iowa) +
  geom_point(mapping = aes(x = yield, y = m_county), alpha = 0.2) + 
  geom_line(mapping = aes(x = yield, y = yield), linetype = "dashed", size=1.5) +
  theme_bw() +
  labs(x = "Actual Yield", y = "Fitted Yield", title= "County Model")
,
ggplot(iowa) +
  geom_point(mapping = aes(x = yield, y = m_combined), alpha = 0.2) + 
  geom_line(mapping = aes(x = yield, y = yield), linetype = "dashed", size=1.5) +
  theme_bw() +
  labs(x = "Actual Yield", y = "Fitted Yield", title= "Combined Model VS Actual", subtitle="Quadratic Temp, Quadratic Year, County")

)


```




### Geospatial trends in corn yields by county across Iowa

There have been significant increases in corn yeild between 1985 & 2018. This increase is observed for all counties with the biggest yields occurring in the east central are of Iowa.

```{r fig.width = 10, fig.height= 8, warning=FALSE}
brks = classInt::classIntervals(
  1:100,
  style = 'fixed',
  fixedBreaks = seq(100,250,25),
  intervalClosure = "right"
)

cornEarly = cornyields %>% 
  filter(year == 1985)
cornLate = cornyields %>% 
  filter(year == 2018) 

mapCornEarly = merge(counties, cornEarly, by.y = "county_name", by.x = "name", all.x=TRUE)
mapCornLate = merge(counties, cornLate, by.y = "county_name", by.x = "name", all.x=TRUE)

```

#### Corn Yields in 1985 vs 2018


```{r, echo=FALSE}

mapview(mapCornEarly, alpha.regions = 0.5, aplha = 1, zcol="yield", 
  at = brks$brks,
  legend = FALSE,
  na.color = '#777777',
  layer.name="Yield (bu/acre)"
) |

mapview(mapCornLate, alpha.regions = 0.5, aplha = 1, zcol="yield",
  at = brks$brks,
  na.color = '#777777',
  layer.name="Yield (bu/acre)"

)


```


## Soybeans

Soybean yield in Iowa see much of the same correlations as corn. Temperature has the strongest relationship with yield, but only up to 48 bushels. After that it is the year, which shows a broad range of impacts but fewer year over year increases than corn. Finally we see that the county have a very narrow band of impact of 10 bushels per acre. An interesting artifact is there seems to be two groups of counties, those above 40 bushels, and those below 40.


```{r, include=FALSE}
## download, format, & clean  NASS data on soybean yields
soyParams <- list(
  commodity_desc = "SOYBEANs", statisticcat_desc="YIELD", 
  prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

soyYields<- nassqs_yields(soyParams)

soyYields$county_ansi <- as.numeric(soyYields$county_ansi)
soyYields$yield <- as.numeric(soyYields$Value)
soyYields$year <- as.numeric(soyYields$year)

# clean and filter this dataset
soyYields <- dplyr::select(soyYields, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))


soy = merge(soyYields, tmaxAll, by.x=c("year", "county_ansi"), by.y=c("year", "countyfp")) %>% 
  mutate(
    yearSq = year^2,
    yieldSq = yield^2,
    meanTmaxSq = meanTmax ^ 2,
    county_name = as.factor(tolower(county_name)),
    county_ansi = as.factor(county_ansi)
  )
```

```{r}
soyMeanTmaxQuadModel<- lm(yield ~ meanTmax + meanTmaxSq, soy)
soy$m_tmax_quad <- soyMeanTmaxQuadModel$fitted.values

soyMeanTmaxLinModel<- lm(yield ~ meanTmax, soy)
soy$m_tmax_lin <- soyMeanTmaxLinModel$fitted.values

soyYearQuadModel<- lm(yield ~ year + yearSq, soy)
soy$m_year_quad <- soyYearQuadModel$fitted.values

soyYearLinModel<- lm(yield ~ year, soy)
soy$m_year_lin <- soyYearLinModel$fitted.values

soyCountyModel<- lm(yield ~ county_name, soy)
soy$m_county <- soyCountyModel$fitted.values

soyCombinedModel<- lm(yield ~ meanTmax + meanTmaxSq + year + yearSq + county_name, soy)
soy$m_combined <- soyCombinedModel$fitted.values

```

```{r, echo=FALSE}
plot_temp = ggplot(soy) +
  geom_point(mapping = aes(x = meanTmax, y = yield), alpha = 0.2, colour = "black") +
  geom_line(mapping = aes(x = meanTmax, y = m_tmax_quad, colour="Quadratic"), size=2) +
  geom_line(mapping = aes(x = meanTmax, y = m_tmax_lin, colour="Linear"), size=2) +
  theme_bw() +
  labs(x = "T Max (°C)", y = "Yield (bsh/ac)", title= "Yield over T Max", color = "Yield Model")

plot_year = ggplot(soy) +
  geom_point(mapping = aes(x = year, y = yield), alpha = 0.2, colour = "black") +
  geom_line(mapping = aes(x = year, y = m_year_quad, colour="Quadratic"), size=2) +
  geom_line(mapping = aes(x = year, y = m_year_lin, colour="Linear"), size=2) +
  theme_bw() +
  labs(x = "Year", y = "Yield (bsh/ac)", title= "Yield over Year", color = "Year Model")

plot_county = ggplot(soy) +
  geom_point(mapping = aes(x = reorder(county_name, -yield), y = yield), alpha = 0.2, colour = "black") +
  geom_line(mapping = aes(x = reorder(county_name, -m_county), y = m_county, colour="Factor"), size=2) +
  theme_bw() +
  labs(x = "County", y = "Yield (bsh/ac)", title= "Yield over County", color = "County Model")

```
```{r, echo=FALSE}

plot_model_temp = ggplot(soy) +
  geom_point(mapping = aes(x = yield, y = m_tmax_quad), alpha = 0.2) +
  geom_line(mapping = aes(x = yield, y = yield), linetype = "dashed", size=2) +
  theme_bw() +
  labs(x = "Actual Yield", y = "Fitted Yield", title= "Quadratic Temp Model", subtitle="Iowa All Years")

plot_model_year = ggplot(soy) +
  geom_point(mapping = aes(x = yield, y = m_year_quad), alpha = 0.2) + 
  geom_line(mapping = aes(x = yield, y = yield), linetype = "dashed", size=2) +
  theme_bw() +
  labs(x = "Actual Yield", y = "Fitted Yield", title= "Quadratic Year Model", subtitle="Iowa All Years")

plot_model_county = ggplot(soy) +
  geom_point(mapping = aes(x = yield, y = m_county), alpha = 0.2) + 
  geom_line(mapping = aes(x = yield, y = yield), linetype = "dashed", size=2) +
  theme_bw() +
  labs(x = "Actual Yield", y = "Fitted Yield", title= "County Model", subtitle="Iowa All Years")

plot_model_combined = ggplot(soy) +
  geom_point(mapping = aes(x = yield, y = m_combined), alpha = 0.2, size=2) + 
  geom_line(mapping = aes(x = yield, y = yield), linetype = "dashed") +
  theme_bw() +
  labs(x = "Actual Yield", y = "Fitted Yield", title= "Combined Model VS Actual", subtitle="Iowa All Years")

```

```{r, fig.cap="Factors of Soybean Yield in Iowa 1980-2018", fig.width = 10, fig.height= 8, res=600, echo=FALSE}


grid.arrange(
top = "Factors of Soybean Yield in Iowa 1980-2018",
layout_matrix = rbind(
  c(1,2,3),
  c(4,5,6),
  c(7)
),
plot_temp, plot_year, plot_county,
plot_model_temp, plot_model_year, plot_model_county,
plot_model_combined
)

```





## Corn VS Soy

This map compares the Iowa corn (left) & soybean (right) yield in 2018. Both legend colors are normalized to the mean. In the west, corn sees lower yields & relative to soy. In the east, corn has higher relative yields to soy.


### Relative Corn VS Soy Yield in Iowa, 2018

```{r, echo=FALSE}
selectYear = 2018

cornTime = cornyields %>% 
  filter(year == selectYear) %>%
  mutate(
    yearSq = as.numeric(year)^2,
    yieldSq = yield^2,
    county_name = as.factor(tolower(county_name)),
    county_ansi = as.factor(county_ansi)
  )

soyTime = soy %>% 
  filter(year == selectYear)


mapCorn = merge(counties, cornTime, by.y = "county_name", by.x = "name", all.x=TRUE)
mapSoy = merge(counties, soyTime, by.y = "county_name", by.x = "name", all.x=TRUE)

mapview(mapCorn, alpha.regions = 0.5, aplha = 1, zcol="yield", 
  legend = TRUE,
  na.color = '#777777',
  layer.name="Corn Yield (bu/acre)"
) |

mapview(mapSoy, alpha.regions = 0.5, aplha = 1, zcol="yield",
  legend = TRUE,
  na.color = '#777777',
  layer.name="Soy Yield (bu/acre)"

)


```


